---
title: "Cat BA Week 1 Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cat BA Week 1 Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(omia)
library(dplyr)
library(ggplot2)
```

## Objective

This vignette demonstrates a practical outbreak workflow using the included
`cat_ba_week1` dataset. The goal is to combine complementary evidence sources:

1. incidence preprocessing and visualization,
2. growth-rate inference from a Poisson model,
3. renewal-equation expected incidence,
4. simulation for uncertainty intuition,
5. optional Rt estimation (when `EpiEstim` is available), and
6. model comparison + consensus summary.

## Data and preprocessing

```{r}
data(cat_ba_week1)
inc <- prep_incidence_data(cat_ba_week1)
inc
```

`prep_incidence_data()` standardizes date/count columns, aggregates duplicate
dates, and fills missing dates. This creates a consistent daily time series for
downstream tools.

## Epidemic curve

```{r fig.height=4, fig.width=7}
plot_epicurve(inc)
```

Visual inspection suggests an upward trajectory with increasing variance, which
is consistent with count-process dynamics and motivates Poisson-family models.

## Growth model (Poisson log-linear)

We fit a Poisson model with a log link:

\[
\log E[I_t] = \beta_0 + \beta_1 t
\]

where `\beta_1` is the daily growth rate. The implied doubling time is
`log(2) / beta1`.

```{r}
growth <- fit_poisson_growth(inc)
growth$estimates
```

```{r fig.height=4, fig.width=7}
growth$fitted |>
  ggplot(aes(date, incidence)) +
  geom_col(fill = "grey80") +
  geom_line(aes(y = fitted), color = "#1f77b4", linewidth = 1) +
  labs(
    title = "Observed incidence vs Poisson growth fit",
    x = NULL,
    y = "Daily cases"
  )
```

## Renewal-equation expected incidence

Given a serial interval distribution `w_s` and Rt, expected incidence is
computed as:

\[
E[I_t] = R_t \sum_{s=1}^{t-1} I_{t-s} w_s
\]

For illustration, use a fixed `Rt = 1.2` and a short serial interval vector.

```{r}
si <- c(0.1, 0.2, 0.3, 0.25, 0.15)
renewal_tbl <- renewal_expected_incidence(
  incidence = inc$incidence,
  rt = 1.2,
  serial_interval = si
)
head(renewal_tbl)
```

## Poisson-process simulation

Simulation helps communicate plausible day-to-day variation around expected
counts.

```{r}
sims <- simulate_poisson_process(
  expected_incidence = renewal_tbl$expected_incidence,
  n_sims = 100,
  seed = 42
)

sim_summary <- sims |>
  group_by(day) |>
  summarise(
    mean_sim = mean(incidence),
    q05 = quantile(incidence, 0.05),
    q95 = quantile(incidence, 0.95),
    .groups = "drop"
  )

head(sim_summary)
```

## Model comparison

A simple practical comparison is between:

- **growth model**: `incidence ~ day` (Poisson)
- **null model**: `incidence ~ 1` (Poisson)

```{r}
model_df <- inc |> mutate(day = as.integer(date - min(date)) + 1)
m_growth <- stats::glm(incidence ~ day, family = poisson(), data = model_df)
m_null <- stats::glm(incidence ~ 1, family = poisson(), data = model_df)

compare_incidence_models(list(growth = m_growth, null = m_null))
```

In this dataset, the growth model should typically outperform the null model on
AIC/BIC, supporting a non-stationary incidence process.

## Consensus summary

```{r}
summary_tbl <- outbreak_consensus_summary(
  incidence_data = inc,
  rt_estimates = tibble::tibble(),
  growth_fit = growth,
  model_comparison = compare_incidence_models(list(growth = m_growth, null = m_null))
)
summary_tbl
```

This summary is designed as a compact handoff table for reports/dashboards.

## Optional Rt estimation with EpiEstim

If `EpiEstim` is installed, integrate time-varying Rt directly:

```{r eval=FALSE}
rt <- estimate_rt_epiestim(inc)
plot_rt_estimates(rt)
```

## Rationale and limitations

### Why combine methods?

- Growth models estimate *trend intensity* and doubling/halving dynamics.
- Renewal calculations encode transmission structure via serial intervals.
- Rt estimation (optional) provides transmissibility over time.
- Model comparison quantifies relative explanatory fit.

No single estimator is sufficient in isolation; agreement across views improves
interpretability.

### Key limitations

- The included dataset is short and intended for demonstration.
- Poisson assumptions may understate overdispersion for real outbreaks.
- Renewal output depends on serial interval specification.
- Rt estimates are method/config-sensitive and should be sensitivity-tested.

For production surveillance, pair this workflow with data quality checks,
multiple model families (e.g., NB/hierarchical variants), and explicit
sensitivity analyses.
