---
title: "Advanced count time-series uncertainty and forecast evaluation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced count time-series uncertainty and forecast evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

```{r setup}
library(omia)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Scope

This vignette demonstrates a reproducible **count time-series** workflow with
explicit uncertainty handling and out-of-sample evaluation. The emphasis is on
methodology:

1. uncertainty estimation at each modeling stage,
2. uncertainty propagation into predictive distributions,
3. interpretation of predictive intervals,
4. rolling-origin backtesting, and
5. forecast scoring (MAE, RMSE, interval coverage, WIS, CRPS).

It also includes a chain-based simulation comparison using `epichains`.

## Data and preprocessing

```{r}
data(cat_ba_week1)
inc <- prep_incidence_data(cat_ba_week1)
inc
```

`prep_incidence_data()` enforces a regular daily index and non-negative counts,
which is essential for coherent rolling-origin resampling and renewal
calculations.

```{r}
plot_epicurve(inc)
```

## Stage 1: growth model + parameter uncertainty

We fit a Poisson log-linear trend model:

\[
\log E[I_t] = \beta_0 + \beta_1 t
\]

with uncertainty in \(\beta_1\) approximated by the model covariance matrix.
From \(\beta_1\), we derive doubling time and an implied reproduction scale via
\(\exp(\beta_1 \times \bar{s})\), where \(\bar{s}\) is mean serial interval.

```{r}
growth <- fit_poisson_growth(
  inc,
  n_draws = 1000,
  mean_si = 4.7,
  seed = 42
)

growth$estimates
```

```{r}
growth_uncertainty <- growth$uncertainty_draws |>
  summarise(
    rt_median = median(implied_rt),
    rt_q10 = quantile(implied_rt, 0.1),
    rt_q90 = quantile(implied_rt, 0.9),
    growth_q10 = quantile(growth_rate, 0.1),
    growth_q90 = quantile(growth_rate, 0.9)
  )

growth_uncertainty
```

The draw object `growth$uncertainty_draws` is reusable in downstream forecasting
and sensitivity analyses.

## Stage 2: renewal forecast with uncertainty propagation

Given serial interval weights \(w_s\), renewal forecasts use

\[
E[I_t] = R_t \sum_{s=1}^{t-1} I_{t-s} w_s
\]

and then sample future counts from a Poisson process. We propagate uncertainty
by using **draw-specific** `R_t` trajectories.

```{r}
serial_interval <- c(0.1, 0.2, 0.3, 0.25, 0.15)
h <- 7

# Map growth-rate uncertainty into Rt paths (constant within horizon per draw)
rt_matrix <- matrix(
  growth$uncertainty_draws$implied_rt,
  nrow = nrow(growth$uncertainty_draws),
  ncol = h
)

forecast_draws <- simulate_renewal_forecast(
  incidence = inc$incidence,
  serial_interval = serial_interval,
  rt = rt_matrix,
  horizon = h,
  seed = 100
)

forecast_summary <- summarise_forecast_uncertainty(
  forecast_draws,
  probs = c(0.1, 0.5, 0.9),
  interval_level = 0.9
)

forecast_summary
```

```{r}
last_date <- max(inc$date)
plot_df <- forecast_summary |>
  mutate(date = last_date + day)

ggplot(plot_df, aes(date, median)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#9ecae1", alpha = 0.45) +
  geom_line(color = "#08519c", linewidth = 0.9) +
  labs(
    title = "7-day predictive distribution from renewal simulation",
    subtitle = "Median and 90% interval with parameter + process uncertainty",
    x = NULL,
    y = "Forecast incidence"
  ) +
  theme_minimal(base_size = 12)
```

### Interpreting predictive uncertainty

A practical interpretation pattern:

- **Median**: central forecast trajectory.
- **90% interval**: expected range under model assumptions.
- **Interval width growth over horizon**: uncertainty accumulation from recursive
  simulation.

To isolate uncertainty sources, compare a fixed-`R_t` run against uncertain
`R_t` draws.

```{r}
fixed_rt_draws <- simulate_renewal_forecast(
  incidence = inc$incidence,
  serial_interval = serial_interval,
  rt = median(growth$uncertainty_draws$implied_rt),
  horizon = h,
  n_sims = 1000,
  seed = 101
)

width_compare <- bind_rows(
  summarise_forecast_uncertainty(fixed_rt_draws, interval_level = 0.9) |>
    transmute(day, width = upper - lower, source = "process only (fixed Rt)"),
  summarise_forecast_uncertainty(forecast_draws, interval_level = 0.9) |>
    transmute(day, width = upper - lower, source = "process + Rt uncertainty")
)

width_compare
```

In this decomposition, wider intervals under stochastic `R_t` indicate
meaningful parameter-driven uncertainty beyond intrinsic count noise.

## Stage 3: rolling-origin backtesting and scoring

Rolling-origin evaluation repeatedly trains on prefixes of the time series and
scores forecasts against unseen future observations.

```{r}
backtest <- rolling_origin_evaluate(
  data = inc,
  initial_window = 14,
  horizon = 5,
  step = 1,
  serial_interval = serial_interval,
  rt_method = "growth",
  n_sims = 400,
  interval_level = 0.9,
  wis_levels = c(0.5, 0.8, 0.95),
  seed = 2026
)

backtest$overall
```

`backtest$overall` provides global summaries across all origins and horizons:

- **MAE**: mean absolute point error,
- **RMSE**: root mean squared point error,
- **coverage**: empirical 90% interval coverage,
- **WIS**: weighted interval score (sharpness + calibration penalties),
- **CRPS**: proper score for full predictive distributions.

```{r}
backtest$origin_summary
```

```{r}
ggplot(backtest$scores, aes(target_date, wis, group = origin_index, color = as.factor(origin_index))) +
  geom_line(alpha = 0.7) +
  labs(
    title = "Rolling-origin WIS by target date",
    x = NULL,
    y = "WIS",
    color = "Origin"
  ) +
  theme_minimal(base_size = 12)
```

## Chain-based uncertainty simulation with epichains

`omia` includes `simulate_epichains_forecast()` as a branching-process
alternative. Here we compare renewal vs epichains forecasts for the same
training origin and scalar `R_t` estimate.

```{r}
# One explicit holdout split for method comparison
train_n <- nrow(inc) - 7
train <- inc[1:train_n, ]
truth <- inc$incidence[(train_n + 1):nrow(inc)]

fit_train <- fit_poisson_growth(train, n_draws = 500, seed = 44)
rt_hat <- median(fit_train$uncertainty_draws$implied_rt)

renewal_cmp <- simulate_renewal_forecast(
  incidence = train$incidence,
  serial_interval = serial_interval,
  rt = rt_hat,
  horizon = 7,
  n_sims = 600,
  seed = 77
)

epichains_cmp <- simulate_epichains_forecast(
  initial_cases = max(1L, round(tail(train$incidence, 1))),
  rt = rt_hat,
  serial_interval = serial_interval,
  horizon = 7,
  n_sims = 600,
  seed = 77
)

score_compare <- bind_rows(
  score_forecast_draws(truth, renewal_cmp) |>
    mutate(method = "renewal_poisson"),
  score_forecast_draws(truth, epichains_cmp) |>
    mutate(method = "epichains_branching")
) |>
  group_by(method) |>
  summarise(
    mae = mean(mae),
    rmse = sqrt(mean(sq_error)),
    coverage = mean(covered),
    wis = mean(wis),
    crps = mean(crps),
    .groups = "drop"
  )

score_compare
```

This side-by-side comparison is useful when teaching how different stochastic
assumptions (renewal-Poisson vs explicit transmission chains) change predictive
sharpness and calibration.

## Summary

The package now supports a complete uncertainty-aware workflow for count
forecasting:

1. estimate trend parameters with uncertainty draws,
2. propagate uncertainty through recursive renewal simulation,
3. summarise predictive distributions,
4. evaluate forecasts with rolling-origin backtesting,
5. report proper scoring rules (WIS, CRPS) alongside MAE/RMSE/coverage,
6. compare against chain-based simulation via `epichains`.

These components are designed for reproducible advanced teaching and practical
forecast model assessment in count time-series settings.
